Hazelnut Live: Type-Theoretic Foundations for Running Incomplete Programs

### 2017-06-02 calculus.tex
bidirectional typing rules
fixes to substitution definition
some tweaks to dynamics, with questions
updating def of err; new question: Do we need both Γ and Δ to judge it?
wrote down major core metatheorems

% \inferrule{ }{
%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
% }
% \end{mathpar}
\caption{Type Consistency and Matched Types}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

@@

\begin{figure}[h!]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] \dehole{\mvar}{\subst}
&=&
\dehole{\mvar}{[\dexp_1 / x] \cdot \subst}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
&=& 
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
\\
{[\dexp_1 / x] \dlam{y}{\dexp_2}}
&=&
{\dlam{y}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dapP{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
{[\dexp_1 / x] \dinj{i}{e_2}}
&=&
{\dinj{i}{[\dexp_1 / x] e_2}}
\\
{[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
&=&
{\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
\\
{[\dexp_1 / x] \dnum{n}}
&=&
{\dnum{n}}
\\
{[\dexp_1 / x] \dadd{e_2}{e_3}}
&=&
{\dadd{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\end{array}
\]
\end{figure}

\begin{figure}
\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
\begin{mathpar}
\Infer{v-lam}
{ }{\isValue{\hlam{x}{\dexp}}}
\and
\Infer{v-num}
{ }{\isValue{\hnum{n}}}
\and
\Infer{v-inj}
{\isValue{\dexp}}{\isValue{\hinj{i}{\dexp}}}
\end{mathpar}

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\Infer{i-ehole}
{ }
{\isIndet{\dehole{\mvar}{\subst}}}
\and
\Infer{i-hole}
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}
\\
\Infer{i-add$_1$}
{\isIndet{\dexp_1} \\ \isFinal{\dexp_2}}
{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{i-add$_2$}
{\isFinal{\dexp_1} \\ \isIndet{\dexp_2}}
{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{i-app}
{\isIndet{\dexp_1}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}
\and
\Infer{i-case}
{\isIndet{\dexp}}
{\isIndet{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
\end{mathpar}

\judgbox{\isErr{\dexp}}{$\dexp$ is error}
\begin{mathpar}
\Infer{e-cast}
% XXX use macro type assignment judgement here
{ \vdash e : \htau_2 \\ \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\dcast{\dexp}{\htau_1}}}
\and
\Infer{e-add$_1$}
{\isErr{\dexp_1}}
{\isErr{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{e-add$_2$}
{\isErr{\dexp_2}}
{\isErr{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{e-app}
{\isErr{\dexp_1}}
{\isErr{\dap{\dexp_1}{\dexp_2}}}
\and
\Infer{e-case}
{\isErr{\dexp}}
{\isErr{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
\end{mathpar}

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\Infer{f-val}
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\Infer{f-indet}
{\isIndet{\dexp}}{\isFinal{\dexp}}
\and
\Infer{f-err}
{\isErr{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
\begin{mathpar}
\Infer{s-hole}
{\stepsTo{\dexp_1}{\dexp_2} }
{\stepsTo{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
\and
\Infer{s-cast}
% XXX use macro type assignment judgement here
{ \vdash e : \htau_2 \\ \tconsistent{\htau_1}{\htau_2}}
{\stepsTo{\dcast{\dexp}{\htau_1}}{\dexp}}
\\
\Infer{s-add$_1$}
{\stepsTo{\dexp_1}{\dexp_1'}}
{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1'}{\dexp_2}}}
\and
\Infer{s-add$_2$}
{\isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1}{\dexp_2'}}}
\and
\Infer{s-add$_3$}
{ n_1 + n_2 = n_3 }
{\stepsTo{\dadd{\dnum{n_1}}{\dnum{n_2}}}{\dnum{n_3}}}
%\and
\\
\Infer{s-app$_1$}
{\stepsTo{\dexp_1}{\dexp_1'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
\and
\Infer{s-app$_2$}
{ \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
\and
\Infer{s-app$_\beta$}
{ \isFinal{\dexp_2} }
{\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
%\and
\\
\Infer{s-case$_1$}
{\stepsTo{\dexp}{\dexp'}}
{\stepsTo{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}{\dcase{\dexp'}{x}{\dexp_x}{y}{\dexp_y}}}
\and
\Infer{s-case$_2$}
{ }
{\stepsTo{\dcase{\dinj{1}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/x]\dexp_x }}
\and
\Infer{s-case$_3$}
{ }
{\stepsTo{\dcase{\dinj{2}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/y]\dexp_y }}
\end{mathpar}
\caption{Structural Dynamics}
\label{fig:isValue}
\label{fig:isIndet}
\label{fig:stepsTo}
\end{figure}

>>>>

% \inferrule{ }{
%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
% }
% \end{mathpar}
\caption{Type Consistency and Matched Types \todo{inconsistent}}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\begin{figure}[h]
\judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
\begin{mathpar}
\inferrule[syn-const]{ }{
  \hsyn{\hGamma}{c}{b}
}

\inferrule[syn-var]{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[syn-alam]{
  \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hsyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[syn-ap]{
  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
}

\inferrule[syn-empty-hole]{ }{
  \hsyn{\hGamma}{\hehole{u}}{\tehole}
}

\inferrule[syn-non-empty-hole]{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}
}

\inferrule[syn-ascription]{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{mathpar}

\judgbox{\hana{\hGamma}{\hexp}{\htau}}{$\htau$ analyzes against type $\htau$}
\begin{mathpar}
\inferrule[ana-lam]{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}

\inferrule[subsumption]{
  \hsyn{\hGamma}{\hexp}{\htau}\\
  \tconsistent{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau'}
}
\end{mathpar}
\caption{Bidirectional Typing of H-Expressions}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

@@

\begin{figure}[h!]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y\\
{[\dexp_1 / x] \dlam{y}{\dexp_2}}
&=&
{\dlam{y}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dap{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
&=& 
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
\end{array}
\]
\end{figure}

\begin{figure}
\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
\begin{mathpar}
\inferrule[v-constant]{ }{
  \isValue{c}
}

\inferrule[v-lam]{ }{
  \isValue{\halam{x}{\htau}{\dexp}}
}
\end{mathpar}

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\inferrule[indet-empty-hole]
{ }
{\isIndet{\dehole{\mvar}{\subst}}}

\inferrule[indet-non-empty-hole]
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[indet-app]
{\isIndet{\dexp_1}\\
% \isFinal{\dexp_2}~\text{\todo{??}}}
 \isFinal{\dexp_2}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}

\inferrule[indet-cast]
{\isIndet{\dexp}}
{\isIndet{\dcast{\dexp}{\htau}}}

\end{mathpar}

\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
\begin{mathpar}
\inferrule[err-nemp-hole~\todo{??}]
{\subst \sim_? \Gamma_2
\\ 
\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[err-cast-error~\todo{??}]
% XXX use macro type assignment judgement here
{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
 \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}

\inferrule[err-lam~\todo{??}]
% XXX use macro type assignment judgement here
{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}

\\

\inferrule[err-app-1]
{
 \isErr{\Delta}{\dexp_1}
 %\\\
 %isFinal{\dexp_2}~\text{\todo{??}}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

\inferrule[err-app-2]
{
 %\isIndet{\dexp_1}
 %\\
 \isErr{\Delta}{\dexp_2}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

\inferrule[err-cast]
% XXX use macro type assignment judgement here
{\isErr{\Delta}{\dexp}}
{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}

\end{mathpar}

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\Infer{f-val}
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\Infer{f-indet}
{\isIndet{\dexp}}{\isFinal{\dexp}}
%% No longer a rule: Errors are not indet anymore:
%% \and
%% \Infer{f-err}
%% {\isErr{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
\begin{mathpar}
\Infer{s-hole}
{\stepsTo{\dexp_1}{\dexp_2} }
{\stepsTo{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
\and
\Infer{s-cast}
% XXX use macro type assignment judgement here
{ \vdash e : \htau_2 \\ \tconsistent{\htau_1}{\htau_2}}
{\stepsTo{\dcast{\dexp}{\htau_1}}{\dexp}}
\\
\Infer{s-app$_1$}
{\stepsTo{\dexp_1}{\dexp_1'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
\and
\Infer{s-app$_2$}
{ \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
\and
\Infer{s-app$_\beta$}
{ \isFinal{\dexp_2} }
{\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
\end{mathpar}
\caption{Structural Dynamics}
\label{fig:isValue}
\label{fig:isIndet}
\label{fig:stepsTo}
\end{figure}

\clearpage
\subsection{Metatheory}

...

\begin{theorem}[Expandability \todo{??}] ~
  \begin{enumerate}
    \item If $\hsyn{\hGamma}{\hexp}{\htau}$ then $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ for some $\dexp$ and $\Delta$.
    \item If $\hana{\hGamma}{\hexp}{\htau}$ then $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ for some $\dexp$ and $\htau'$ and $\Delta$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Correspondence \todo{??}] ~
  \begin{enumerate}
    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hsyn{\hGamma}{\hexp}{\htau}$.
    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\hana{\hGamma}{\hexp}{\htau}$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Typed Expansion] ~
  \begin{enumerate}
    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$ then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Expansion Unicity] ~
  \begin{enumerate}
    \item If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ and $\expandSyn{\hGamma}{\hexp}{\htau'}{\dexp'}{\Delta'}$ then $\htau=\htau'$ and $\dexp=\dexp'$ and $\Delta=\Delta'$.
    \item If $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp}{\htau_2}{\Delta}$ and $\expandAna{\hGamma}{\hexp}{\htau_1}{\dexp'}{\htau_2'}{\Delta'}$ then $\dexp=\dexp'$ and $\htau_2=\htau_2'$ and $\Delta=\Delta'$.
  \end{enumerate}
\end{theorem}

\begin{theorem}[Type Assignment Unicity]
If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$ then $\htau=\htau'$.
\end{theorem}

\begin{theorem}[Progress]
If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either $\isValue{\dexp}$ or $\isIndet{\dexp}$ or $\isErr{\Delta}{\dexp}$ or $\stepsTo{\dexp}{\dexp'}$.
\end{theorem}

\begin{theorem}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and $\stepsTo{\dexp}{\dexp'}$ then $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$. \todo{or maybe up to consistency?}
\end{theorem}

\todo{complete progress + preservation}

\todo{canonical forms}

\todo{indeterminate forms}
